# Seamless phase-II/III study

```{r}
#| echo: false
#| warning: false

library(ggplot2)
library(tidyverse)
library(splitstackshape)
library(extraDistr)
```

This vignette was motivated by a consulting of CTU Bern (C2597). It is an example of a Bayesian seamless phase-II/III design with a correlated adverse event endpoint and a time-to-event primary endpoint.

## Available evidence

For the **control arm** we used information from a published randomized controlled trial. The authors reported from approximately 350 total randomized patients to a control arm a disease-free survival percentage of approximately 50% at 24-months. For our sample size calculation we assumed a slightly more conservative percentage of 45%.

For the **experimental arm** we use information from another study with a similar intervention which reported an approximate 70% progression-free survival at 24 months. We assumed a more conservative 60% progression-free survival at 24-months in the experimental arm.

For **adverse events outcomes** (defined by grade 3 CTCAE) in the experimental arm we used the reported number of a review from 20 (4 prospective studies and 16 retrospective studies). For the prospective studies the authors reported a weighted average of 13.9% adverse events, whereas for the retrospective studies a value of 10.4%. We used a slightly more conservative value of 15% for the sample size calculation.

No assumptions for the correlation between adverse events and disease-free survival were reported. In our used simulation approach we assumed correlations of 0.1, 0.5 and 0.9 between the adverse event outcome and the progression-free survival.

## Endpoints and testing hierarchy

**Safety endpoint**: Adverse events in experimental arm (for phase-II part)

**Primary endpoint**: Disease-free survival at 24-months (phase-III endpoint)

Of note, these are not co-primary endpoints, because the safety endpoint is not required to define a trial success. But we test the endpoints in a hierarchical way: Only if the probability for an excess adverse event proportion is below a given threshold, the primary endpoint will be assessed for futility assessment.

## Hypotheses

We consider the following one-sided hypotheses for the phase-III trial

- Null hypothesis: Log hazard ratio $\theta$ between experimental arm and control arm is greater or equal than zero, that is, $H_0: \theta \geq 0$,
- Alternative hypothesis: Log hazard ratio $\theta$ between experimental arm and control arm is smaller than zero, that is, $H_0: \theta<0$.

## Trial structure

```{mermaid}
gantt
    title Seamless Phase-II/III design
    dateFormat X
    axisFormat %s
    section Phase-II
    Interim phase-II: active, 0, 50
    Phase-II: 0, 100
    Evaluation for Phase-III: crit, 0, 100
    section Phase-III
    Phase-III: 0, 360
```  


## Methods

### Gaussian approximation

We approximate the log hazard ratio of disease-free survival by
$$
\theta=log(HR)=\frac{log(s_1)}{log(s_0)}
$$
where $s_1$ is the disease-free survival probability of the experimental arm and $s_0$ is the disease-free survival probability of the control arm. In our case we have $s_1=0.6$ and $s_0=0.45$ leading to a $log(HR)=0.64$.

We approximate the variance of $\theta$ with $4/m$, where $m$ is the number of events (see @Spiegelhalter2003, chapter 2.4.2).

### Phase-II part

We plan the phase-II part with an a priori fixed sample size of 100 patients. A safety interim analysis is planned after 25 patients are recruited in the experimental arm.

### Early stopping phase-II

We plan an safety interim analysis which assess' the probability for an excess adverse event proportion using a Bayesian predictive probability in the **experimental arm**.

The trial is stopped early if $PP>\theta_{S}$, for an upper discontinuation proportion threshold $p_{max}$,
$$
\begin{aligned}
PP&=E\left[I\left\{ P\left(p_1>p_{max}|r_{1}, n_{1}, s_{1}\right)>\theta_{T_{Non-safe}}\right\}|r_{1}, n_{1}\right] \\
&=\sum_{s_1=0}^{m_1} I(P(p_1 > p_{max}|r_{1},n_1,s_1)>\theta_{T_{Non-safe}}) P(S_{1}=s_1|r_{1}, n_1),
\end{aligned}
$$
where $r_1$ (number of observed adverse events at interim analysis), $n_1$ (number of patients at interim analysis), $s_1$ (expected future events at interim analysis) and a threshold $\theta_{T_{Non-safe}}$.

If at this stage the trial is considered as non-safe we stop the trial. Otherwise we recruit more patients until the planned sample size is reached.


### Decision for phase-III

The decision to continue seamlessly into a phase-III trial is based on two criteria

- 1) We stop the trial at the end of phase-II part if there is a high probability for excess adverse events (same as described above).
- 2) If we have a low predictive probability for a successful phase-III trial at the end of phase-II.

For the latter approach we use a Bayesian predictive 'interim monitoring' approach as described in @Spiegelhalter2003, chapter 6.6.3. In brief, we use the observed data at the end of phase-II (interim data $y_m$), the prior information and the assumed number of 'future' phase-III patients ($Y_n$) to calculate the predictive probability of $P(\theta<0|y_m, Y_n)$. We stop the trial for futility if $P(\theta<0|y_m, Y_n)<\epsilon_{fut}$, which is

$$
1-\Phi\left[ \frac{\sqrt{m_{phase3}}\theta_{phase3}}{2}+\frac{m_{phase2}\theta_{phase2}}{2\sqrt{m_{phase3}}}+\frac{m_{prior}\theta_{prior}}{2\sqrt{m_{phase3}}}+\sqrt{\frac{m_{prior}+m_{phase2}+m_{phase3}}{m_{phase3}}}\cdot z_{1-\epsilon_{fut}} \right],
$$
where $z_{1-\epsilon_{fut}}$ is the upper $1-\epsilon_{fut}$%-quantile of a Gaussian distribution.

The trial at the end of phase-III is considered as successful if the posterior distribution of the log hazard ratio $P(\theta<0|\theta_0, y_m, Y)\geq 1-\epsilon_{success}$.


### Simulation assumptions

We assumed bivariate exponentially distributed time to event outcomes for adverse events and disease-free survival. We used the following componentwise rate parameters

| Parameter                                                            | Value   |
|----------------------------------------------------------------------|---------|
| Exponential rate parameter for time to disease progression or death (control arm) | 1/30      |
| Exponential rate parameter for time to disease progression or death (experimental arm) | 1/47     |
| Exponential rate parameter for adverse events (experimental arm) | 1/37      |

for simulating data. Based on these parameters we get the proportion of disease progression or death at 24-months in the control arm of

```{r}
#| echo: true
#| warning: false

rate_event_0 <- 1/30

# Proportion of patients with disease progression or death in control arm at 24-months 
paste0(round(pexp(24, rate_event_0)*100, 0), "%")
```

and in the experimental arm of
```{r}
#| echo: true
#| warning: false

rate_event_1 <- 1/47

# Proportion of patients with disease progression or death in experimental arm at 24-months 
paste0(round(pexp(24, rate_event_1)*100, 0), "%")
```

and the proportion of adverse events at 6-months in the experimental arm of

```{r}
#| echo: true
#| warning: false

rate_tox_1 <- 1/47

# Proportion of patients with disease progression or death in experimental arm at 6-months 
paste0(round(pexp(6, rate_tox_1)*100, 0), "%")
```

For the sample size calculation we assumed that for patients with an adverse event, 24-month progression-free survival information will be available and used for the sample size calculation.

#### Stopping boundaries phase-II

We assumed the following parameters for the excess adverse event proportion stopping in the experimental arm:

| Parameter                                                            | Value   |
|----------------------------------------------------------------------|---------|
| Excess threshold | 25%      |
| Beta prior parameters | a=0.2, b=0.8     |
| Initial interim sample size | n=25      |
| Final sample size | n=50      |
| $\theta_{T_{Non-safe}}$ | 60%      |
| $\theta_{S}$ | 60%      |

This gives us the following stopping boundaries based on the predictive probability:

```{r}
#| echo: true
#| warning: false

p_max <- c(0.25)

# Prior information
a_1 <- 0.2
b_1 <- 0.8

x <- seq(0,1,0.001)

n_initial <- 25
r <- 0:n_initial
n_increase <- 25

n_max <- 50
n <- seq(n_initial, n_max, n_increase)
r <- 0:n_max
m <- n_max-r

#### Scenario
theta_T <- 0.6



data <- expand.grid(n=n, r=r) %>% arrange(n) %>% 
  filter(n>=r) %>% mutate(i=n_max-n+1)
data <- expandRows(data, count=3)
data <- data %>% group_by(n, r) %>% 
  mutate(m=n_max-n, ind=1, i=cumsum(ind)-1, ind=NULL)

data <- data %>% 
  mutate(PS1=dbbinom(i, size=m, alpha=a_1+r, beta=b_1+n-r), 
                        Ti=1-pbeta(p_max, a_1+r+i, b_1+n_max-r-i), 
                        ind=ifelse(Ti>theta_T, 1, 0), n_max)

data <- data %>% select(n_max, n, r, m, i, Ti, PS1, ind)

threshold_safety <- 0.6

data_pp <- data %>% group_by(n_max, n, r) %>% summarise(pp=round(sum(PS1*ind),6), 
                                                 stop_pp=ifelse(pp>threshold_safety, 1, 0))


stop_boundaries0 <- data_pp %>% filter(stop_pp==1) %>%
  group_by(n_max, n) %>% summarise(r=min(r), stop=NULL)

names(stop_boundaries0) <- c("Maximal sample size", "Interim sample size", "Stopping boundary")
stop_boundaries0
```

### Priors

We use the following priors for the simulation

#### Beta prior for excess adverse event stopping

As mentioned in the section of 'available evidence' there is some evidence for the adverse event proportion in the experimental arm. We assumed a slighty more conservative mean value of 0.2 with a prior weight of 1 patient from a beta prior with a=0.2 and b=0.8.

#### Gaussian prior on hazard ratio for disease-free survival

```{r}
#| echo: true
#| warning: false

rate_event_0 <- 1/30
rate_event_1 <- 1/47
expected_event_rate <- 0.5*(pexp(24, rate_event_0)+pexp(24, rate_event_1))

prior_mean <- log(0.8)
prior_weight <- 90

prior_sd <- 2/sqrt(prior_weight*expected_event_rate)
```

For the prior on the hazard ratio for disease-free survival we assumed a Gaussian distribution centered on a mean of $log(0.8)$ with a standard deviation of $`r round(prior_sd, 2)`$, which is based on the expected event rate with a prior weight of $`r prior_weight`$ patients. This leads to a probability that the prior log hazard ratio is greater than 0 of $`r round(1-pnorm(0, prior_mean, prior_sd),2)`$.

```{r}
#| echo: true
#| warning: false

x <- seq(-1.5, 1, 0.001)

data_plot <- tibble(x=x, y=dnorm(x, prior_mean, prior_sd))
# data_plot$x <- exp(data_plot$x)
# data_plot$y <- exp(data_plot$y)

ggplot(data_plot, aes(x, y))+geom_line()+theme_bw()+xlab("Log hazard ratio")+ylab("Density")+geom_vline(xintercept = 0, linetype="dashed")+theme(panel.grid.minor = element_blank())+ggtitle("Prior on hazard ratio")
```

## Results

We simulated 1,000 trials with the above specification parameters and report the following trial operating characteristics:

| Parameter                                                            | Value  |
|----------------------------------------------------------------------|---------|
| Probability of stopping at interim analysis (phase-II) because of safety under null hypothesis |  5.8%  |
| Probability of stopping (phase-II) because of safety under null hypothesis | 15.4%  |
| Probability of starting a phase-III trial under null hypothesis | 43.1%  |
| Probability of not-starting a phase-III trial because of futility under null hypothesis | 27.7%  |
| Probability of successful phase-III trial under the null hypothesis |     3%  |
| Probability of successful phase-III trial under the alternative hypothesis |  81%     |
| $\epsilon_{fut}$ |  10%     |
| $\epsilon_{success}$ |  2.5%     |


## Comparison with frequentist phase-III trial

We compare our seamless phase-II/III design simulation results with a sample size calculation from a 'classic' frequentist phase-III design with a one-sided logrank test ($\alpha=0.025$, $1-\beta=0.8$) using the *artsurv* command in Stata:


**artsurv**, method(l) edf0(0.45) hr(1,0.64) np(2) recrt(0 0, 1, 0 ) onesided(1) alpha(0.025) power(0.8)


This command gives a calculated total sample size of 336 patients with an expected total number of events of 160 events.

## Author {-}

André Moser, Senior Statistician\
Iulia-Maia Muresan, Statistician\
Department of Clinical Research\
University of Bern\
3012 Bern, Switzerland