# Two-arm study with a binary primary outcome for assessing threshold stopping

```{r}
#| echo: false
#| warning: false

library(ggplot2)
library(tidyverse)
library(splitstackshape)
library(extraDistr)
```

This vignette was motivated by a study of CTU Bern (1938). It is an example of a Bayesian phase-II trial design with interim threshold stopping using predictive probabilities (@Lee2008, @Sambucini2021). The primary endpoint is a binary endpoint, say, treatment discontinuation (TD).

The trial is designed with early stopping decisions rules based on predictive probabilities that the intervention is non-safe, that is, a high probability of an excess TD proportion given interim data. We quantify the trial operations characteristics probability of stopping because of non-safety (stopping trial before maximal sample size is reached), probability of treatment discontinuation and the expected number of patients in experimental arm using a simulation approach.

## Methods

### Sample size

The sample size was fixed to $N_{max}=36$ patients (24 patients in the experimental arm, 12 patients in the control arm) because of feasibility.

### Study design

The following figure summarises the trial design. Suppose we randomise patients with a 2:1 allocation ratio to the experimental arm and the control arm. Let $p_i$ be the TD proportion coming from a design prior $\pi^d_i \sim Beta(a^d_i,b^d_i)$, $p_i$, $i=0,1$. The design prior represents the uncertainty of $p_i$ at the design stage and might be different from the analysis prior. One interim analysis will be performed after the recruitment of $N^{I}=n_{0}+n_{1}=18$ patients ($n_{0}$ denotes the number of patients in the control arm and $n_{1}$ denotes the number of patients in the experimental arm; The index $I$ stands for 'interim') which corresponds to 50% of the fixed sample size.

```{mermaid}
flowchart TD
  A[2:1 randomisation] --> B[Experimental group with TD proportion \np1 coming from design prior]
  A --> C[Control group with TD proportion \np0 coming from design prior]
  B ~~~ D[Interim analysis\n\nStop early:\nIf experimental group has a high predictive probability of excessive discontinuation\notherwise randomise new patients until N_max is reached]
  C ~~~ D
  style A fill:white,color:black
  style B fill:white,color:black
  style C fill:white,color:black
  style D fill:white,color:black,stroke-dasharray: 5 5
```

### Stopping criterium

The excessive threshold for stopping the trial was set to $p_{max}=0.2$. This number was based on the design prior choice (see subsection 'Design prior choice'). In brief, based on a discussion with clinical experts and available evidence the assumed mean average TD proportion was set to $0.1$. A doubling of this number led to $p_{max}=0.2$. 

### Statistical methods

Suppose that at the interim analysis **in the experimental arm** $r_{1}$ discontinuation events are observed. If we assume that $N_{max}=n_{max, 0}+n_{max, 1}$ is the planned sample size of the trial, then a remaining future $m_{1}=n_{max, 1}-n_{1}$ will be recruited in the experimental arm with possible future events $S_1\in\{0,\cdots, m_1\}$. Note that at each interim analysis the posterior distribution is Beta-distributed $p_i|r_i,n_i \sim Beta(a_i+r_i, b_i+n_i-r_i)$ and $S_1|r_1,n_1$ is Beta-binomial distributed $S_1 \sim Betabinomal(m_1, a+r_1, b+n_1-r_1)$.

The trial is stopped early if $PP>\theta_{S}$, for an upper discontinuation proportion threshold $p_{max}$,
$$
\begin{aligned}
PP&=E\left[I\left\{ P\left(p_1>p_{max}|r_{1}, n_{1}, s_{1}\right)>\theta_{T_{Non-safe}}\right\}|r_{1}, n_{1}\right] \\
&=\sum_{s_1=0}^{m_1} I(P(p_1 > p_{max}|r_{1},n_1,s_1)>\theta_{T_{Non-safe}}) P(S_{1}=s_1|r_{1}, n_1).
\end{aligned}
$$
That is, we stop the trial early if there is a high predictive probability $\Theta_S$ that the discontinuation proportion in the experimental arm exceeds a upper threshold $p_{max}$ given interim data and future events, otherwise the remaining 18 patients are randomised until the fixed sample size is reached.

Stopping boundaries are calculated for each combination of $r_1$ (and thus $m_1$) given the maximal sample size $N_{max}$ and the interim sample size $N^I$. The tibble below shows the example of $r_1=2$ and $n_1=12$ for the skeptical prior. We set $\theta_{T_{Non-safe}}=0.6$.

```{r}
#| echo: true
#| warning: false

p_max <- c(0.2)

# Prior information
prior_type <- "Skeptical"
a_1 <- 2.4
b_1 <- 9.6

x <- seq(0,1,0.001)

n_initial <- 12
r <- 0:n_initial
n_increase <- 12

n_max <- 24
n <- seq(n_initial, n_max, n_increase)
r <- 0:n_max
m <- n_max-r

#### Scenario
theta_T <- 0.6



data <- expand.grid(n=n, r=r) %>% arrange(n) %>% 
  filter(n>=r) %>% mutate(i=n_max-n+1)
data <- expandRows(data, count=3)
data <- data %>% group_by(n, r) %>% 
  mutate(m=n_max-n, ind=1, i=cumsum(ind)-1, ind=NULL)

data <- data %>% 
  mutate(PS1=dbbinom(i, size=m, alpha=a_1+r, beta=b_1+n-r), 
                        Ti=1-pbeta(p_max, a_1+r+i, b_1+n_max-r-i), 
                        ind=ifelse(Ti>theta_T, 1, 0), n_max)

data <- data %>% select(n_max, n, r, m, i, Ti, PS1, ind)
data %>% filter(n==12, r==2)
```

The predictive probabilities for each $r_1$ are (with $\theta_{S}=0.8$):

```{r}
#| echo: true
#| warning: false


## Early stopping using predictive distribution

threshold_safety <- 0.8

data_pp <- data %>% group_by(n_max, n, r) %>% summarise(pp=round(sum(PS1*ind),6), 
                                                 stop_pp=ifelse(pp>threshold_safety, 1, 0))
data_pp %>% filter(n==12)
```

The same calculation holds for the neutral prior and we get the following stopping boundaries

```{r}
#| echo: true
#| warning: false

stop_boundaries0 <- data_pp %>% filter(stop_pp==1) %>%
  group_by(n_max, n) %>% summarise(r=min(r), stop=NULL)
stop_boundaries0$p_max <- p_max
stop_boundaries0$theta_T <- theta_T
stop_boundaries0$threshold_safety <- threshold_safety
stop_boundaries0$prior_type <- prior_type
stop_boundaries <- stop_boundaries0

# Prior information
prior_type <- "Neutral"
a_1 <- 0.6
b_1 <- 5.4

x <- seq(0,1,0.001)

n_initial <- 12
r <- 0:n_initial
n_increase <- 12

n_max <- 24
n <- seq(n_initial, n_max, n_increase)
r <- 0:n_max
m <- n_max-r

#### Scenario
theta_T <- 0.6
threshold_safety <- 0.8


data <- expand.grid(n=n, r=r) %>% arrange(n) %>% filter(n>=r) %>% mutate(i=n_max-n+1)
data <- expandRows(data, count=3)
data <- data %>% group_by(n, r) %>% mutate(m=n_max-n, ind=1, i=cumsum(ind)-1, ind=NULL)

data <- data %>% mutate(cond_postprob=dbbinom(i, size=m, alpha=a_1+r, beta=b_1+n-r), 
                        Ti=1-pbeta(p_max, a_1+r+i, b_1+n_max-r-i), 
                        ind=ifelse(Ti>theta_T, 1, 0), n_max)

data <- data %>% select(n_max, n, r, m, i, cond_postprob, Ti, ind)



## Early stopping using predictive distribution
data_pp <- data %>% group_by(n_max, n, r) %>% summarise(pp=round(sum(cond_postprob*ind),6), 
                                                 stop_pp=ifelse(pp>threshold_safety, 1, 0))



stop_boundaries0 <- data_pp %>% filter(stop_pp==1) %>%
  group_by(n_max, n) %>% summarise(r=min(r), stop=NULL)
stop_boundaries0$p_max <- p_max
stop_boundaries0$theta_T <- theta_T
stop_boundaries0$threshold_safety <- threshold_safety
stop_boundaries0$prior_type <- prior_type
stop_boundaries <- bind_rows(stop_boundaries, stop_boundaries0)


stop_boundaries %>% filter(n==12)
```


#### Design prior choice

The design prior for the primary endpoint was chosen as based on clinical expert knowledge and available evidence. We assumed that the "true" TD proportion comes from a $Beta(1.2,10.8)$ design prior which is centered at a TD proportion of 10% with a 90% uncertainty range from 0.9% to 26.6%. The probability that the discontinuation rate is larger than 20% is 12.2%. The design prior has weight of 12 patients, that is, half of planned patients in the experimental arm.

```{r}
#| echo: false
#| warning: false

# Prior information control
a_0 <- 1.2
b_0 <- 10.8

x <- seq(0.001,0.999,0.001)

# qbeta(0.05, a_0, b_0)
# qbeta(0.95, a_0, b_0)
# pbeta(0.03, a_0, b_0)
# 1-pbeta(0.15, a_0, b_0)
# 1-pbeta(0.20, a_0, b_0)

data_prior <- data.frame(x, y=dbeta(x, a_0, b_0), type=1)

# Prior information experimental
# a_0 <- 5/4
# b_0 <- 45/4


data_prior$type <- factor(data_prior$type, levels=1, labels=c("a=1.2, b=10.8 (both arms)"))
fig <- ggplot(data_prior, aes(x=x, y=y, linetype=type, group=type))+
  geom_line()+theme_bw()+scale_x_continuous(breaks=seq(0,1,0.1))+
  theme(panel.grid = element_blank(), legend.position = "bottom")+
  scale_linetype("Prior parameters")+
  geom_vline(xintercept=0.1, colour="red", linetype=2)+
  ylab("Density")+xlab("")+ggtitle("Beta design prior")+
  labs(caption=c("Dashed red vertical lines indicates mean value of design prior."))
fig
```

#### Analysis prior choice

As analysis priors we decided for two scenarios.

- A skeptical Beta(2.4, 9.4) analysis prior for the experimental arm which is centered on a TD proportion of 20% with a probability of excessing 40% of $5.6$%.
- A neutral Beta(0.6, 5.4) analysis prior for the experimental arm which is centered on a discontinuation proportion of 10%. 

```{r}
#| echo: false
#| warning: false

# Skeptical prior
a_0 <- 2.4
b_0 <- 9.6

x <- seq(0.001,0.999,0.001)

data_prior <- data.frame(x, y=dbeta(x, a_0, b_0), type=1)

# Neutral prior
a_0 <- 0.6
b_0 <- 5.4

data_prior <- bind_rows(data_prior, data.frame(x, y=dbeta(x, a_0, b_0), type=2))


data_prior$type <- factor(data_prior$type, levels=1:2, 
                          labels=c("a=2.4, b=9.6 (Skeptical prior)", "a=0.6, b=5.4 (Neutral prior)"))
fig <- ggplot(data_prior %>% filter(y<5), aes(x=x, y=y, linetype=type, group=type))+
  geom_line()+theme_bw()+scale_x_continuous(breaks=seq(0,1,0.1))+
  theme(panel.grid = element_blank(), legend.position = "bottom", legend.direction = "vertical")+
  scale_linetype("Prior parameters")+geom_vline(xintercept=0.2, colour="red", linetype=1)+
  geom_vline(xintercept=0.1, colour="red", linetype=2)+
  ylab("Density")+xlab("")+ggtitle("Beta analysis prior")+
  labs(caption=c("Red vertical lines indicate mean values of priors."))
fig
```


### Simulation study

We set up simulation study to quantify operations characteristics

- Probability of stopping because of non-safety (stopping trial before $N_{max}$ is reached),
- Probability of treatment discontinuation (declaring treatment discontinuation including $N_{max}$),
- Expected number of patients in experimental arm

with the following parameters

- Beta design priors: Boths arms: a=1.5, b=13.5. True $p_i$ are drawn from the design priors,
- Beta analysis priors: Boths arms: Skeptical: a=2.4, b=9.6. Neutral: a=0.6, b=5.4,
- Maximal discontinuation proportion threshold $p_{max}$: 0.2,
- Number of randomised patients in the experimental arm at first interim analysis $n_1$: 12,
- Number of randomised patients in the control arm at first interim analysis $n_1$: 6,
- Excessive discontinuation probability for posterior distribution $\theta_{T}$: 0.6,
- Safety threshold for predictive distribution $\theta_{S}$: 0.8,
- Number of simulation runs: 10'000,

- Maximal sample size $N_{max}$: 12/24 patients per arm,
- Increase of sample size $n_{increase}$: 6/12 patients per arm.

## Results

The results from our simulation study are:

| Operations characteristics                                  | $N_{max}=36$   |
|-------------------------------------------------------------|----------------|
| Probability of intolerable trial (skeptical prior)          | 15.8%            |
| Probability of intolerable trial (neutral prior)            | 8.2%            |
| Probability of early stopping because of non-safety         | 7.7%           |
| Expected number of patients in experimental arm             | 23.1           |

| Stopping boundaries                                         |    |
|-------------------------------------------------------------|----------------|
| Stopping boundary interim analysis (skeptical prior)        | 4+           |
| Intolerable trial at trial end (skeptical prior)      | 6+           |
| Intolerable trial at trial end (neutral prior)        | 7+           |

## Author {-}

André Moser, Senior Statistician\
Department of Clinical Research\
University of Bern\
3012 Bern, Switzerland